---
title: "miaTime: Microbiome Time Series Analysis"
date: "`r Sys.Date()`"
author:
    - name: Leo Lahti
    email: leo.lahti@iki.fi
    - name: Yagmur Simsek
    email: yagmur.simsek@hsrw.org
package: 
    miaTime
output: 
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: yes
        toc_depth: 2
        number_sections: true
vignette: >
    %\VignetteIndexEntry{miaTime}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                        fig.width = 9,
                        message = FALSE,
                        warning = FALSE)
```

# Introduction

`miaTime` implements tools for microbiome time series manipulation based on the
`SummarizedExperiment` [@SE] 

# Installation

To install `miaTime`, install `devtools` first, if it is not installed.
Afterwards use the `install_github` function from `devtools`.

```{r, eval=FALSE}
library(devtools)
devtools::install_github("microbiome/miaTime")
```

# Load *miaTime*

```{r load-packages, message=FALSE, warning=FALSE}
library("miaTime")
```

# Functions

`TimePointsToDate` converts time points to POSIXct class time object. 
The dataset stored in `miaTime` can be used as an example.
```{r}
library(miaTime)
data(hitchip1006)
force(hitchip1006)
HitchipDate <- TimePointsToDate(hitchip1006, "time", origin = "1970-01-01")
```

`addOrderSequence` adds a new field that gives the rank of the chosen field.
```{r}
data(GlobalPatterns, package="mia")
se <- GlobalPatterns
se1 <- addOrderSequence(se,field = "SampleType",
            rank_field_name = "SampleType_rank",
            na.last = TRUE, ties.method = "first")
```

The functions in `miaTime` are applied on `SummarizedExperiment`
object or `TreeSummarizedExperiment` object. It returns same object 
class as given.

# Session info

```{r}
sessionInfo()
```
