---
title: "miaTime: Microbiome Time Series Analysis"
date: "`r Sys.Date()`"
author:
- name: Leo Lahti
  email: leo.lahti@iki.fi
- name: Yagmur Simsek
  email: yagmur.simsek@hsrw.org
package: 
    miaTime
output: 
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: yes
        toc_depth: 2
        number_sections: true
vignette: >
    %\VignetteIndexEntry{miaTime}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r eval=FALSE, include=FALSE}
library(Cairo)
knitr::opts_chunk$set(cache = FALSE,
                      fig.width = 9,
                      dpi=300,
                      dev = "png",
                      dev.args = list(type = "cairo-png"),
                      message = FALSE,
                      warning = FALSE)
```

```{r loadhide, message=FALSE, warning=FALSE, echo=FALSE}
library(mia)
library(miaTime)
library(dplyr)
library(lubridate)
library(SummarizedExperiment)
```

# Introduction

`miaTime` implements tools for time series manipulation based on the `TreeSummarizedExperiment` [@TSE] data container. Much of the functionality is also applicable to the `SummarizedExperiment` [@SE] data objects.

This tutorial shows how to analyse divergence, convergence, and
dynamics in time series.


## Installation

Installing the latest development version in R.

```{r, eval=FALSE}
library(devtools)
devtools::install_github("microbiome/miaTime")
```

Loading the package:

```{r load-packages, message=FALSE, warning=FALSE}
library(miaTime)
```



## Estimating divergence and convergence

Calculate divergence from a baseline sample (or another reference sample).
This calculation can be carried out with the  function \link[mia]{estimateDivergence}. The `mia` package can be installed with the following lines. 

```{r eval=FALSE, include=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("mia")
```


Calculate the difference in time from the baseline time point per
subject. The first (smallest) time point is used as the baseline by
default.

```{r}
library(mia)   # Load the functions
library(miaTime)   # Load the functions
set.seed(41287)    # Set random seed
data(hitchip1006)  # Example data
tse <- hitchip1006 # Rename data object

# Subset to speed up this example
# Just pick two example subjects that have 5 time points each
selected.samples <- sample(names(which(sort(table(colData(hitchip1006)$subject))==5)), 2)
tse <- tse[, colData(tse)$subject %in% selected.samples]

# Add relative abundances
tse <- transformSamples(tse, abund_values="counts", method="relabundance")

# Calculate divergence from baseline sample (time point 1)
library(dplyr)
tse <- getBaselineDivergence(tse,
                              group = "subject",
                              time_field = "time",
                              name_divergence = "divergence_from_baseline",
                              name_timedifference = "time_from_baseline",
                              abund_values="relabundance",
                              FUN = vegan::vegdist,
                              method="bray")

# Remove the baseline time points
df <- as.data.frame(colData(tse)) %>% dplyr::filter(time_from_baseline > 0)

# Visualize time difference vs. community divergence (dissimilarity)
library(ggplot2)
theme_set(theme_bw(20))
ggplot(data=df, aes(x=time_from_baseline, y=divergence_from_baseline, group=subject)) +
    geom_point(aes(color=subject)) +
    geom_smooth(formula = y ~ x, method="lm")
```



Define the baseline manually for each subject.

```{r}
# First define the function that defines the baseline for a given data object:
base <- function (tse) {
  baseline_sample <- rownames(colData(tse)[which.min(colData(tse)$time),])
  baseline_sample
}

# List samples per subject
spl <- split(colnames(tse), droplevels(tse$subject))

# Find the baseline per subject
baselines <- sapply(spl, function (s) {base(tse[,s])})

# Calculate divergence from baseline sample (time point 1)
library(dplyr)
tse <- getBaselineDivergence(tse,
                              group = "subject",
                              time_field = "time",
                              name_divergence = "divergence_from_baseline",
                              name_timedifference = "time_from_baseline",
                              abund_values="relabundance",
                              FUN = vegan::vegdist,
                              method="bray",
			      baseline_sample=baselines)

# Remove the baseline time points
df <- as.data.frame(colData(tse)) %>% dplyr::filter(time_from_baseline > 0)

# Visualize time difference vs. community divergence (dissimilarity)
library(ggplot2)
theme_set(theme_bw(20))
ggplot(data=df, aes(x=time_from_baseline, y=divergence_from_baseline, group=subject)) +
    geom_point(aes(color=subject)) +
    geom_smooth(formula = y ~ x, method="lm")
```


Similar example without miaTime. First without subject-wise grouping:


```{r}
# Identify the baseline sample
tse <- hitchip1006
baseline_sample <- colnames(tse[, which.min(colData(tse)$time)])

# Pick the reference sample from count assay and calculate euclidean distance
# Name the new field as "divergence_from_baseline".
# This contains the divergence information.
library(mia)
baseline_profile <- assays(tse)$counts[, baseline_sample]
print(length(baseline_profile))
print(dim(tse))
tse <- estimateDivergence(tse, name = "divergence_from_baseline", 
                         reference = baseline_profile,
                         FUN = stats::dist, method = "euclidean")

# Also add time from baseline
colData(tse)$time_from_baseline <- colData(tse)[, "time"] -
                                  colData(tse)[baseline_sample, "time"]

# Visualize time difference vs. community divergence (dissimilarity)
library(dplyr)
df <- as.data.frame(colData(tse)) %>% dplyr::filter(time_from_baseline > 0)

library(ggplot2)
ggplot(data=df, aes(x=time_from_baseline, y=divergence_from_baseline)) +
    geom_point() +
    geom_smooth(formula = y ~ x, method="lm")
```


Divergence from baseline per subject:

```{r, eval=FALSE}
# First define the function that calculates divergence for a given SE object
# in a desired way:
fun <- function (tse) {
  baseline_sample <- rownames(colData(tse)[which.min(colData(tse)$time),])
  estimateDivergence(tse, name = "divergence_from_baseline", 
                     reference = assays(tse)$counts[, baseline_sample],
		                 FUN = stats::dist, method = "euclidean")

}

# List samples per subject
spl <- split(colnames(tse), tse$subject)

# Apply the operation per subject (here just the first 3 to speed up example)
tse_list <- lapply(spl[1:3], function (s) {fun(tse[,s])})

# Combine the TSE elements in the list
tse2 <- do.call(cbind, tse_list)
```

Calculate convergence i.e. divergence from the last sample:

```{r}
# Identify the last sample
tse <- hitchip1006
endpoint_sample <- colnames(tse[, which.max(colData(tse)$time)])

# Pick the reference sample from count assay and calculate euclidean distance
# Name the new field as "convergence".
# This contains the divergence information.
library(mia)
tse <- estimateDivergence(tse, name = "convergence", 
                         reference = assays(tse)$counts[, endpoint_sample], 
                         FUN = stats::dist, method = "euclidean")

# Also add time from endpoint
colData(tse)$time_from_endpoint <- colData(tse)[, "time"] -
                                  colData(tse)[endpoint_sample, "time"]

```

Sample dissimilarity between consecutive time steps(step size n >= 1) within
a group(subject, age, reaction chamber, etc.) can be calculated by `getStepwiseDivergence`.

```{r}
library(miaTime)
data("hitchip1006")
tse <- hitchip1006

tse <- tse[, colData(tse)$subject %in% c("900", "934", "843", "875")]
tse2 <- getStepwiseDivergence(tse, group = "subject",
                            time_interval = 1,
                            time_field = "time")

#for multiple time steps in a different grouping
tse3 <- getStepwiseDivergence(tse, group = "age",
                           time_interval = 2,
                           time_field = "time")
```


## Example Minimal gut microbiome

Dense samples of the minimal gut microbiome. In the initial hours, MDb-MM was grown under batch condition and 24 h onwards, continuous feeding of media with pulse feeding cycles. This information is stored in the `colData`.   

```{r fig.width=8, fig.height=3}
library(miaTime)
data("minimalgut")
tse <- minimalgut
# quick check of number of samples 
table(colData(tse)$StudyIdentifier,colData(tse)$condition_1)

```

Visualize samples available for each of the bioreactors. This allows to identify if there are any missing samples for specific times.  

```{r}
library(ggplot2)
# Visualize sampling times available in the experiment 
colData(tse) |> 
    as.data.frame() |> 
    ggplot(aes(as.factor(Time.hr), StudyIdentifier)) +
    geom_tile(aes(fill=condition_1), color="white") +
    scale_fill_manual("Condition Sampled", 
                      values = c("#ff006e", "#e07a5f", "#457b9d")) +
    theme_minimal() +
    theme(axis.text.x = element_text(size=8, angle = 90),
          legend.position = "top") +
    labs(x="Time (h)", y="")
```


## Community dynamics   

The `minimalgut` dataset, mucus-diet based minimal microbiome
(MDbMM-16), consists of 16 species assembled in three bioreactors. We
can investigate the succession of mdbMM16 from the start of experiment
here hour zero until the end of the experiment.

```{r}
## Divergence from baseline i.e from hour zero.
tse <- mia::relAbundanceCounts(minimalgut) # get relative abundance
tse <- getBaselineDivergence(tse,
                             group = "StudyIdentifier",
                             time_field = "Time.hr",
                             name_divergence = "divergence_from_baseline",
                             name_timedifference = "time_from_baseline",
                             abund_values="relabundance",
                             FUN = vegan::vegdist,
                             method="bray")

```


Visualize the divergence

```{r fig.height=4, fig.width=8}
# First define nice colors for bioreactors
bioreac_cols <- c(`Bioreactor A`= "#b2182b",
                  `Bioreactor B`="#2166ac",
                  `Bioreactor C` = "#35978f")

colData(tse) |>
    as.data.frame() |>
    ggplot(aes(x=Time.hr, y=divergence_from_baseline))+
    geom_point(aes(color=StudyIdentifier), size=2, alpha=0.5) +
    geom_line(aes(color=StudyIdentifier)) +
    theme_minimal() +
    scale_color_manual(values = bioreac_cols) +
    labs(x="Time (h)", y="Divergence \nfrom baseline") +
    # highlight specific timepoints
    geom_vline(xintercept = 152, lty=2, color="#991720") + 
    geom_vline(xintercept = 248, lty=2, color= "#0963bd")+
    annotate("text",x=c(152, 248),y=c(0.8, 0.8),
             label=c("Addition of\nB.hydrogenotrophica","Acetate Discontinued"),
             hjust=c(1.05,-0.05))
```

Now visualize abundance of *Blautia hydrogenotrophica* using the `miaViz::plotSeries` function.

```{r fig.height=4, fig.width=8}
library(miaViz)
plotSeries(mia::relAbundanceCounts(minimalgut),
           x = "Time.hr",
           y = "Blautia_hydrogenotrophica",
           colour_by = "Species",
           abund_values = "relabundance")+
    geom_vline(xintercept = 152, lty=2, color="#991720") + 
    geom_vline(xintercept = 248, lty=2, color= "#0963bd")+
    annotate("text",x=c(152, 248),y=c(0.2, 0.15),
             label=c("Addition of\nB.hydrogenotrophica","Acetate Discontinued"),
             hjust=c(1.05,-0.05))+
    labs(x="Time (h)", y="B.hydrogenotrophica\nRelative Abundance") +
    theme(legend.position = "none")
    
```


# Session info

```{r}
sessionInfo()
```
