---
title: "miaTime: Microbiome Time Series Analysis"
date: "`r Sys.Date()`"
author:
- name: Leo Lahti
  email: leo.lahti@iki.fi
- name: Yagmur Simsek
  email: yagmur.simsek@hsrw.org
package: 
    miaTime
output: 
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: yes
        toc_depth: 2
        number_sections: true
vignette: >
    %\VignetteIndexEntry{miaTime}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r eval=FALSE, include=FALSE}
library(Cairo)
knitr::opts_chunk$set(cache = FALSE,
                      fig.width = 9,
                      dpi=300,
                      dev = "png",
                      dev.args = list(type = "cairo-png"),
                      message = FALSE,
                      warning = FALSE)
```




# Introduction

`miaTime` implements tools for microbiome time series manipulation based
on the `TreeSummarizedExperiment` [@TSE] data container. Much of the functionality is also applicable to the `SummarizedExperiment` [@SE] data objects.


## Installation

Installing the latest development version in R.

```{r, eval=FALSE}
library(devtools)
devtools::install_github("microbiome/miaTime")
```

Loading the package:

```{r load-packages, message=FALSE, warning=FALSE}
library("miaTime")
```


## Storing time information with Period class

`miaTime` utilizes the functions available in the package `lubridate`
to convert time series field to "Period" class object. This gives access to a 
number of readily available [time series manipulation tools](https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html).


Load example data:

```{r}
# Load packages
library(miaTime)
library(lubridate)
library(SummarizedExperiment)

# Load demo data
data(hitchip1006)
tse <- hitchip1006

# Time is given in days in the demo data.
# Convert days to seconds
time_in_seconds <- 60*60*24*colData(tse)[,"time"]
# Convert the time data to period class
Seconds <- as.period(time_in_seconds, unit="sec")
# Check the output
Seconds[1140:1151]
```


## Conversion between time units

The time field in days is now shown in seconds. It can then be
converted to many different units using the lubridate package.

```{r}
Hours <- as.period(Seconds, unit = "hour")
Hours[1140:1151]
```

The updated time information can then be added to the
`SummarizedExperiment` data object as a new `colData` (sample data)
field.


```{r}
colData(tse)$timeSec <- Seconds
colData(tse)
```

## Calculating time differences 

The \link[lubridate]{as.duration} function helps to specify time points as durations.

```{r}
Duration <- as.duration(Seconds)
Duration[1140:1151]
```

The difference between subsequent time points can then be calculated.

```{r}
Timediff <- diff(Duration)
Timediff <- c(NA, Timediff)
Timediff[1140:1151]
```

The time difference from a selected point to the other time points
can be calculated as follows.

```{r}
base <- Hours - Hours[1] #distance from starting point
base[1140:1151]

base_1140 <- Seconds - Seconds[1140]
base_1140[1140:1151]
```

## Time point rank

Rank of the time points can be calculated by `rank` function provided in base R.

```{r}
colData(tse)$rank <- rank(colData(tse)$time)
colData(tse)
```


## Operations per unit

Sometimes we need to operate on time series per unit (subject,
reaction chamber, sampling location, ...).

Add time point rank per subject.

```{r}
library(dplyr)
colData(tse) <- colData(tse) %>%
   as.data.frame() %>%
   group_by(subject) %>%
   mutate(rank = rank(time, ties.method="average")) %>%
   DataFrame()
```

Calculate the difference in time from a given baseline time point.

```{r}
# Define baseline function;
base <- function (x, baseline=min(x)) {x - baseline}
filter <- dplyr::filter

# Use time 1 as the baseline
colData(tse) <- colData(tse) %>%
    as.data.frame() %>%
    group_by(subject) %>%
    mutate(shifted = base(time, baseline=1)) %>%
    DataFrame()
# Check output
as.data.frame(colData(tse)[, c("subject", "time", "shifted")]) %>%
    filter(subject == "831")

# Use the minimum time point per subject as the baseline:
colData(tse) <- colData(tse) %>%
    as.data.frame() %>%
    group_by(subject) %>%
    mutate(shifted = base(time)) %>%
    DataFrame()

# Check output
as.data.frame(colData(tse)[, c("subject", "time", "shifted")]) %>%
    filter(subject == "831")
```

## Subset to baseline samples

```{r}
#samples with time point 0
se <- hitchip1006
sample_time_zero <- colData(se)[which(colData(se)[, "time"] == 0),]

#sample with the smallest time point within each subject
colData(se) <- colData(se) %>%
   as.data.frame() %>%
   group_by(subject) %>%
   mutate(rank = rank(time, ties.method="average")) %>%
   DataFrame()

time_within_subject <- colData(se)[which(colData(se)$rank == 1),]

# Add some NAs in the time point data to demonstrate functionality with missing data
nainds <- sample(seq_len(ncol(se)), 50)
colData(se)$time[nainds] <- NA

colData(se) <- colData(se) %>%
   as.data.frame() %>%
   group_by(subject) %>%
   mutate(rank = rank(time, ties.method="average")) %>%
   DataFrame()

time_within_subject <- colData(se)[which(colData(se)$rank == 1),]

#There is still one sample from each subject even if the time information of the sample is missing
check_time <- data.frame(time_within_subject) %>% select(subject, time, rank)

```


## Estimating divergence and convergence

Calculate divergence from a baseline sample (or another reference sample).
This calculation can be carried out with the  function \link[mia]{estimateDivergence}. The `mia` package can be installed with the following lines. 

```{r eval=FALSE, include=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("mia")
```


Calculate the difference in time from the baseline (first) time point, per subject.

```{r}
# Subset to speed up example
library(mia)
tse <- hitchip1006
set.seed(41287)
# Just pick two example subjects that have 5 time points each
selected.samples <- sample(names(which(sort(table(colData(hitchip1006)$subject))==5)), 2)
tse <- tse[, colData(tse)$subject %in% selected.samples]
# Add relative abundances
tse <- transformSamples(tse, abund_values="counts", method="relabundance")
# Calculate divergence from baseline sample (time point 1)
tse <- getBaselineDivergence(tse,
                              group = "subject",
                              time_field = "time",
                              name_divergence = "divergence_from_baseline",
                              name_timedifference = "time_from_baseline",
                              abund_values="relabundance",
                              FUN = vegan::vegdist,
                              method="bray")

# Visualize time difference vs. community divergence (dissimilarity)
library(dplyr)

# Remove the baseline time points
df <- as.data.frame(colData(tse)) %>% dplyr::filter(time_from_baseline > 0)

library(ggplot2)
theme_set(theme_bw(20))
ggplot(data=df, aes(x=time_from_baseline, y=divergence_from_baseline, group=subject)) +
    geom_point(aes(color=subject)) +
    geom_smooth(formula = y ~ x, method="lm")
```



Similar example without miaTime. First without subject-wise grouping:

```{r}
# Identify the baseline sample
tse <- hitchip1006
baseline_sample <- colnames(tse[, which.min(colData(tse)$time)])

# Pick the reference sample from count assay and calculate euclidean distance
# Name the new field as "divergence_from_baseline".
# This contains the divergence information.
library(mia)
baseline_profile <- assays(tse)$counts[, baseline_sample]
print(length(baseline_profile))
print(dim(tse))
tse <- estimateDivergence(tse, name = "divergence_from_baseline", 
                         reference = baseline_profile,
                         FUN = stats::dist, method = "euclidean")

# Also add time from baseline
colData(tse)$time_from_baseline <- colData(tse)[, "time"] -
                                  colData(tse)[baseline_sample, "time"]

# Visualize time difference vs. community divergence (dissimilarity)
library(dplyr)
df <- as.data.frame(colData(tse)) %>% dplyr::filter(time_from_baseline > 0)

library(ggplot2)
ggplot(data=df, aes(x=time_from_baseline, y=divergence_from_baseline)) +
    geom_point() +
    geom_smooth(formula = y ~ x, method="lm")
```


Divergence from baseline per subject:

```{r, eval=FALSE}
# First define the function that calculates divergence for a given SE object
# in a desired way:
fun <- function (tse) {
  baseline_sample <- rownames(colData(tse)[which.min(colData(tse)$time),])
  estimateDivergence(tse, name = "divergence_from_baseline", 
                     reference = assays(tse)$counts[, baseline_sample],
		                 FUN = stats::dist, method = "euclidean")

}

# List samples per subject
spl <- split(colnames(tse), tse$subject)

# Apply the operation per subject (here just the first 3 to speed up example)
tse_list <- lapply(spl[1:3], function (s) {fun(tse[,s])})

# Combine the TSE elements in the list
tse2 <- do.call(cbind, tse_list)
```

Calculate convergence i.e. divergence from the last sample:

```{r}
# Identify the last sample
tse <- hitchip1006
endpoint_sample <- colnames(tse[, which.max(colData(tse)$time)])

# Pick the reference sample from count assay and calculate euclidean distance
# Name the new field as "convergence".
# This contains the divergence information.
library(mia)
tse <- estimateDivergence(tse, name = "convergence", 
                         reference = assays(tse)$counts[, endpoint_sample], 
                         FUN = stats::dist, method = "euclidean")

# Also add time from endpoint
colData(tse)$time_from_endpoint <- colData(tse)[, "time"] -
                                  colData(tse)[endpoint_sample, "time"]

```

Sample dissimilarity between consecutive time steps(step size n >= 1) within
a group(subject, age, reaction chamber, etc.) can be calculated by `getStepwiseDivergence`.

```{r}
library(miaTime)
data("hitchip1006")
tse <- hitchip1006

tse <- tse[, colData(tse)$subject %in% c("900", "934", "843", "875")]
tse2 <- getStepwiseDivergence(tse, group = "subject",
                            time_interval = 1,
                            time_field = "time")

#for multiple time steps in a different grouping
tse3 <- getStepwiseDivergence(tse, group = "age",
                           time_interval = 2,
                           time_field = "time")
```


## Example Minimal gut microbiome

Dense samples of the minimal gut microbiome. In the initial hours, MDb-MM was grown under batch condition and 24 h onwards, continuous feeding of media with pulse feeding cycles. This information is stored in the `colData`.   

```{r fig.width=8, fig.height=3}
library(miaTime)
data("minimalgut")
tse <- minimalgut
# quick check of number of samples 
table(colData(tse)$StudyIdentifier,colData(tse)$condition_1)

```

Visualize samples available for each of the bioreactors. This allows to identify if there are any missing samples for specific times.  

```{r}
library(ggplot2)
# Visualize sampling times available in the experiment 
colData(tse) |> 
    as.data.frame() |> 
    ggplot(aes(as.factor(Time.hr), StudyIdentifier)) +
    geom_tile(aes(fill=condition_1), color="white") +
    scale_fill_manual("Condition Sampled", 
                      values = c("#ff006e", "#e07a5f", "#457b9d")) +
    theme_minimal() +
    theme(axis.text.x = element_text(size=8, angle = 90),
          legend.position = "top") +
    labs(x="Time (h)", y="")
```


## Community dynamics   

The `minimalgut` dataset, mucus-diet based minimal microbiome
(MDbMM-16), consists of 16 species assembled in three bioreactors. We
can investigate the succession of mdbMM16 from the start of experiment
here hour zero until the end of the experiment.

```{r}
## Divergence from baseline i.e from hour zero.
tse <- mia::relAbundanceCounts(minimalgut) # get relative abundance
tse <- getBaselineDivergence(tse,
                             group = "StudyIdentifier",
                             time_field = "Time.hr",
                             name_divergence = "divergence_from_baseline",
                             name_timedifference = "time_from_baseline",
                             abund_values="relabundance",
                             FUN = vegan::vegdist,
                             method="bray")

```


Visualize the divergence

```{r fig.height=4, fig.width=8}
# First define nice colors for bioreactors
bioreac_cols <- c(`Bioreactor A`= "#b2182b",
                  `Bioreactor B`="#2166ac",
                  `Bioreactor C` = "#35978f")

colData(tse) |>
    as.data.frame() |>
    ggplot(aes(x=Time.hr, y=divergence_from_baseline))+
    geom_point(aes(color=StudyIdentifier), size=2, alpha=0.5) +
    geom_line(aes(color=StudyIdentifier)) +
    theme_minimal() +
    scale_color_manual(values = bioreac_cols) +
    labs(x="Time (h)", y="Divergence \nfrom baseline") +
    # highlight specific timepoints
    geom_vline(xintercept = 152, lty=2, color="#991720") + 
    geom_vline(xintercept = 248, lty=2, color= "#0963bd")+
    annotate("text",x=c(152, 248),y=c(0.8, 0.8),
             label=c("Addition of\nB.hydrogenotrophica","Acetate Discontinued"),
             hjust=c(1.05,-0.05))
```

Now visualize abundance of *Blautia hydrogenotrophica* using the `miaViz::plotSeries` function.

```{r fig.height=4, fig.width=8}
library(miaViz)
plotSeries(mia::relAbundanceCounts(minimalgut),
           x = "Time.hr",
           y = "Blautia_hydrogenotrophica",
           colour_by = "Species",
           abund_values = "relabundance")+
    geom_vline(xintercept = 152, lty=2, color="#991720") + 
    geom_vline(xintercept = 248, lty=2, color= "#0963bd")+
    annotate("text",x=c(152, 248),y=c(0.2, 0.15),
             label=c("Addition of\nB.hydrogenotrophica","Acetate Discontinued"),
             hjust=c(1.05,-0.05))+
    labs(x="Time (h)", y="B.hydrogenotrophica\nRelative Abundance") +
    theme(legend.position = "none")
    
```


# Session info

```{r}
sessionInfo()
```
