---
title: "miaTime: Microbiome Time Series Analysis"
date: "`r Sys.Date()`"
author:
- name: Leo Lahti
  email: leo.lahti@iki.fi
- name: Yagmur Simsek
  email: yagmur.simsek@hsrw.org
package: 
    miaTime
output: 
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: yes
        toc_depth: 2
        number_sections: true
vignette: >
    %\VignetteIndexEntry{miaTime}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                        fig.width = 9,
                        message = FALSE,
                        warning = FALSE)
```

# Introduction

`miaTime` implements tools for microbiome time series manipulation based on the
`SummarizedExperiment` [@SE] 

# Installation

To install `miaTime`, install `devtools` first, if it is not installed.
Afterwards use the `install_github` function from `devtools`.

```{r, eval=FALSE}
library(devtools)
devtools::install_github("microbiome/miaTime")
```

# Load *miaTime*

```{r load-packages, message=FALSE, warning=FALSE}
library("miaTime")
```

# Functions

`miaTime` utilizes the functions available in the package 
`lubridate` to convert time series field to "Period" class object.

The dataset `hitchip1006` stored in `miaTime` can be used as an example.
```{r}
library(miaTime)
library(lubridate)
library(SummarizedExperiment)
data(hitchip1006)

Seconds <- as.period(60*60*24*colData(hitchip1006)[,"time"], unit="sec")
Seconds[1140:1151]
```

The time field in days is now converted to seconds. It can then be converted to 
different units.

```{r}
Hours <- as.period(Seconds, unit = "hour")
Hours[1140:1151]
```
This information can then be added to the SummarizedExperiment object 
as new fields.

```{r}
colData(hitchip1006)$timeSec <- Seconds
colData(hitchip1006)
```

The `as.duration` function helps to specify time points as durations. 
Durations are stored in seconds, the only time unit with a consistent length.
```{r}
Duration <- as.duration(Seconds)
Duration[1140:1151]
```

The difference between subsequent time points can then be calculated.

```{r}
Timediff <- diff(Duration)
Timediff <- c(NA, Timediff)
Timediff[1140:1151]
```

Also, the time difference from selected point to rest of each time point 
can be calculated.
```{r}
base <- Hours - Hours[1] #distance from starting point
base[1140:1151]

base_1140 <- Seconds - Seconds[1140]
base_1140[1140:1151]

```
Rank of the time points can be calculated by `rank` function provided in base R.
```{r}
colData(hitchip1006)$rank <- rank(colData(hitchip1006)$time)
colData(hitchip1006)
```

# Session info

```{r}
sessionInfo()
```
