---
title: "miaTime: Microbiome Time Series Analysis"
date: "`r Sys.Date()`"
author:
- name: Leo Lahti
  email: leo.lahti@iki.fi
- name: Yagmur Simsek
  email: yagmur.simsek@hsrw.org
package: 
    miaTime
output: 
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: yes
        toc_depth: 2
        number_sections: true
vignette: >
    %\VignetteIndexEntry{miaTime}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, echo=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                        fig.width = 9,
                        message = FALSE,
                        warning = FALSE)
```

# Introduction

`miaTime` implements tools for microbiome time series manipulation based
on the `SummarizedExperiment` [@SE] data container.


## Installation

Installing the latest development version in R.

```{r, eval=FALSE}
library(devtools)
devtools::install_github("microbiome/miaTime")
```

Loading the package:

```{r load-packages, message=FALSE, warning=FALSE}
library("miaTime")
```


## Storing time information with Period class

`miaTime` utilizes the functions available in the package `lubridate`
to convert time series field to "Period" class object. This gives access to a 
number of readily available [time series manipulation tools](https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html).


Load example data:

```{r}
# Load packages
library(miaTime)
library(lubridate)
library(SummarizedExperiment)

# Load demo data
data(hitchip1006)
se <- hitchip1006

# Time is given in days in the demo data.
# Convert days to seconds
time_in_seconds <- 60*60*24*colData(se)[,"time"]
# Convert the time data to period class
Seconds <- as.period(time_in_seconds, unit="sec")
# Check the output
Seconds[1140:1151]
```


## Conversion between time units

The time field in days is now shown in seconds. It can then be
converted to many different units using the lubridate package.

```{r}
Hours <- as.period(Seconds, unit = "hour")
Hours[1140:1151]
```

The updated time information can then be added to the
`SummarizedExperiment` data object as a new `colData` (sample data)
field.


```{r}
colData(se)$timeSec <- Seconds
colData(se)
```

## Calculating time differences 

The \link[lubridate]{as.duration} function helps to specify time points as durations.

```{r}
Duration <- as.duration(Seconds)
Duration[1140:1151]
```

The difference between subsequent time points can then be calculated.

```{r}
Timediff <- diff(Duration)
Timediff <- c(NA, Timediff)
Timediff[1140:1151]
```

The time difference from a selected point to the other time points
can be calculated as follows.

```{r}
base <- Hours - Hours[1] #distance from starting point
base[1140:1151]

base_1140 <- Seconds - Seconds[1140]
base_1140[1140:1151]
```

## Time point rank

Rank of the time points can be calculated by `rank` function provided in base R.

```{r}
colData(se)$rank <- rank(colData(se)$time)
colData(se)
```


## Operations per unit

Sometimes we need to operate on time series per unit (subject,
reaction chamber, sampling location, ...).

Add time point rank per subject.

```{r}
library(dplyr)
colData(se) <- colData(se) %>%
   as.data.frame() %>%
   group_by(subject) %>%
   mutate(rank = rank(time, ties.method="average")) %>%
   DataFrame()
```

Calculate the difference in time from a given baseline time point.

```{r}
# Define baseline function;
base <- function (x, baseline=min(x)) {x - baseline}
filter <- dplyr::filter

# Use time 1 as the baseline
colData(se) <- colData(se) %>%
    as.data.frame() %>%
    group_by(subject) %>%
    mutate(shifted = base(time, baseline=1)) %>%
    DataFrame()
# Check output
as.data.frame(colData(se)[, c("subject", "time", "shifted")]) %>%
    filter(subject == "831")

# Use the minimum time point per subject as the baseline:
colData(se) <- colData(se) %>%
    as.data.frame() %>%
    group_by(subject) %>%
    mutate(shifted = base(time)) %>%
    DataFrame()

# Check output
as.data.frame(colData(se)[, c("subject", "time", "shifted")]) %>%
    filter(subject == "831")
```

## Subset to baseline samples

```{r}
#samples with time point 0
se <- hitchip1006
sample_time_zero <- colData(se)[which(colData(se)[, "time"] == 0),]

#sample with the smallest time point within each subject
colData(se) <- colData(se) %>%
   as.data.frame() %>%
   group_by(subject) %>%
   mutate(rank = rank(time, ties.method="average")) %>%
   DataFrame()

time_within_subject <- colData(se)[which(colData(se)$rank == 1),]

# Add some NAs in the time point data to demonstrate functionality with missing data
nainds <- sample(seq_len(ncol(se)), 50)
colData(se)$time[nainds] <- NA

colData(se) <- colData(se) %>%
   as.data.frame() %>%
   group_by(subject) %>%
   mutate(rank = rank(time, ties.method="average")) %>%
   DataFrame()

time_within_subject <- colData(se)[which(colData(se)$rank == 1),]

#There is still one sample from each subject even if the time information of the sample is missing
check_time <- data.frame(time_within_subject) %>% select(subject, time, rank)

```


## Estimating divergence and convergence

Calculate divergence from a baseline sample (or another reference sample).
This calculation can be carried out with the  function \link[mia]{estimateDivergence}. The `mia` package can be installed with the following lines. 

```{r eval=FALSE, include=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("mia")
```


```{r}
# Identify the baseline sample
se <- hitchip1006
baseline_sample <- colnames(se[, which.min(colData(se)$time)])

# Pick the reference sample from count assay and calculate euclidean distance
# Name the new field as "divergence_from_baseline".
# This contains the divergence information.
library(mia)
baseline_profile <- assays(se)$counts[, baseline_sample]
print(length(baseline_profile))
print(dim(se))
se <- estimateDivergence(se, name = "divergence_from_baseline", 
                         reference = baseline_profile,
                         FUN = stats::dist, method = "euclidean")

# Also add time from baseline
colData(se)$time_from_baseline <- colData(se)[, "time"] -
                                  colData(se)[baseline_sample, "time"]

# Visualize time difference vs. community divergence (dissimilarity)
library(dplyr)
df <- as.data.frame(colData(se)) %>% dplyr::filter(time_from_baseline > 0)

library(ggplot2)
ggplot(data=df, aes(x=time_from_baseline, y=divergence_from_baseline)) +
    geom_point() +
    geom_smooth(formula = y ~ x, method="lm")
```


Divergence from baseline per subject:

```{r, eval=FALSE}
# First define the function that calculates divergence for a given SE object
# in a desired way:
fun <- function (se) {
  baseline_sample <- rownames(colData(se)[which.min(colData(se)$time),])
  estimateDivergence(se, name = "divergence_from_baseline", 
                     reference = assays(se)$counts[, baseline_sample],
		                 FUN = stats::dist, method = "euclidean")

}

# List samples per subject
spl <- split(colnames(se), se$subject)

# Apply the operation per subject (here just the first 3 to speed up example)
se_list <- lapply(spl[1:3], function (s) {fun(se[,s])})

# Combine the SE elements in the list
se2 <- do.call(cbind, se_list)
```

Calculate convergence i.e. divergence from the last sample:

```{r}
# Identify the last sample
se <- hitchip1006
endpoint_sample <- colnames(se[, which.max(colData(se)$time)])

# Pick the reference sample from count assay and calculate euclidean distance
# Name the new field as "convergence".
# This contains the divergence information.
library(mia)
se <- estimateDivergence(se, name = "convergence", 
                         reference = assays(se)$counts[, endpoint_sample], 
                         FUN = stats::dist, method = "euclidean")

# Also add time from endpoint
colData(se)$time_from_endpoint <- colData(se)[, "time"] -
                                  colData(se)[endpoint_sample, "time"]

```

Sample dissimilarity between consecutive time steps(step size n >= 1) within
a group(subject, age, reaction chamber, etc.) can be calculated by `getTimeDivergence`.

```{r}
library(miaTime)
data("hitchip1006")
se <- hitchip1006

se <- se[, colData(se)$subject %in% c("900", "934", "843", "875")]
se2 <- getTimeDivergence(se, group = "subject",
                            time_interval = 1,
                            time_field = "time")

#for multiple time steps in a different grouping
se3 <- getTimeDivergence(se, group = "age",
                           time_interval = 2,
                           time_field = "time")
```

# Session info

```{r}
sessionInfo()
```
