---
title: "Time series manipulation"
date: "`r Sys.Date()`"
author:
- name: Leo Lahti
  email: leo.lahti@iki.fi
- name: Yagmur Simsek
  email: yagmur.simsek@hsrw.org
package: 
    miaTime
output: 
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: yes
        toc_depth: 2
        number_sections: true
vignette: >
    %\VignetteIndexEntry{miaTime}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r eval=FALSE, include=FALSE}
library(Cairo)
knitr::opts_chunk$set(cache = FALSE,
                      fig.width = 9,
                      dpi=300,
                      dev = "png",
                      dev.args = list(type = "cairo-png"),
                      message = FALSE,
                      warning = FALSE)
```

```{r loadhide, message=FALSE, warning=FALSE, echo=FALSE}
library(mia)
library(miaTime)
library(dplyr)
library(lubridate)
library(SummarizedExperiment)
```

# Introduction

`miaTime` implements tools for time series manipulation based on the `TreeSummarizedExperiment` [@TSE] data container. Much of the functionality is also applicable to the `SummarizedExperiment` [@SE] data objects. This tutorial shows how to use `miaTime` methods as well as the broader R/Bioconductor ecosystem to manipulate time series data.


## Installation

Installing the latest development version in R.

```{r, eval=FALSE}
library(devtools)
devtools::install_github("microbiome/miaTime")
```

Loading the package:

```{r load-packages, message=FALSE, warning=FALSE}
library(miaTime)
```


## Sorting samples 

The _tidySingleCellExperiment_ package provides handy functions for (Tree)SE manipulation.

```{r sort samples, message=FALSE, warning=FALSE}
library("tidySingleCellExperiment")
data("hitchip1006")
tse <- hitchip1006
tse2 <- tse %>% tidySingleCellExperiment::arrange(subject, time)
```


## Storing time information with Period class

`miaTime` utilizes the functions available in the package `lubridate`
to convert time series field to "Period" class object. This gives access to a 
number of readily available [time series manipulation tools](https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html).


Load example data:

```{r}
# Load packages
library(miaTime)
library(lubridate)
library(SummarizedExperiment)

# Load demo data
data(hitchip1006)
tse <- hitchip1006

# Time is given in days in the demo data.
# Convert days to seconds
time_in_seconds <- 60*60*24*colData(tse)[,"time"]
# Convert the time data to period class
Seconds <- as.period(time_in_seconds, unit="sec")
# Check the output
Seconds[1140:1151]
```


## Conversion between time units

The time field in days is now shown in seconds. It can then be
converted to many different units using the lubridate package.

```{r}
Hours <- as.period(Seconds, unit = "hour")
Hours[1140:1151]
```

The updated time information can then be added to the
`SummarizedExperiment` data object as a new `colData` (sample data)
field.


```{r}
colData(tse)$timeSec <- Seconds
colData(tse)
```

## Calculating time differences 

The \link[lubridate]{as.duration} function helps to specify time points as durations.

```{r}
Duration <- as.duration(Seconds)
Duration[1140:1151]
```

The difference between subsequent time points can then be calculated.

```{r}
Timediff <- diff(Duration)
Timediff <- c(NA, Timediff)
Timediff[1140:1151]
```

The time difference from a selected point to the other time points
can be calculated as follows.

```{r}
base <- Hours - Hours[1] #distance from starting point
base[1140:1151]

base_1140 <- Seconds - Seconds[1140]
base_1140[1140:1151]
```

## Time point rank

Rank of the time points can be calculated by `rank` function provided in base R.

```{r}
colData(tse)$rank <- rank(colData(tse)$time)
colData(tse)
```


## Operations per unit

Sometimes we need to operate on time series per unit (subject,
reaction chamber, sampling location, ...).

Add time point rank per subject.

```{r}
library(dplyr)
colData(tse) <- colData(tse) %>%
   as.data.frame() %>%
   group_by(subject) %>%
   mutate(rank = rank(time, ties.method="average")) %>%
   DataFrame()
```

Calculate the difference in time from a given baseline time point.

```{r}
# Define baseline function;
base <- function (x, baseline=min(x)) {x - baseline}
filter <- dplyr::filter

# Use time 1 as the baseline
colData(tse) <- colData(tse) %>%
    as.data.frame() %>%
    group_by(subject) %>%
    mutate(shifted = base(time, baseline=1)) %>%
    DataFrame()
# Check output
as.data.frame(colData(tse)[, c("subject", "time", "shifted")]) %>%
    filter(subject == "831")

# Use the minimum time point per subject as the baseline:
colData(tse) <- colData(tse) %>%
    as.data.frame() %>%
    group_by(subject) %>%
    mutate(shifted = base(time)) %>%
    DataFrame()

# Check output
as.data.frame(colData(tse)[, c("subject", "time", "shifted")]) %>%
    filter(subject == "831")
```

## Subset to baseline samples

```{r}
#samples with time point 0
se <- hitchip1006
sample_time_zero <- colData(se)[which(colData(se)[, "time"] == 0),]

#sample with the smallest time point within each subject
colData(se) <- colData(se) %>%
   as.data.frame() %>%
   group_by(subject) %>%
   mutate(rank = rank(time, ties.method="average")) %>%
   DataFrame()

time_within_subject <- colData(se)[which(colData(se)$rank == 1),]

# Add some NAs in the time point data to demonstrate functionality with missing data
nainds <- sample(seq_len(ncol(se)), 50)
colData(se)$time[nainds] <- NA

colData(se) <- colData(se) %>%
   as.data.frame() %>%
   group_by(subject) %>%
   mutate(rank = rank(time, ties.method="average")) %>%
   DataFrame()

time_within_subject <- colData(se)[which(colData(se)$rank == 1),]

#There is still one sample from each subject even if the time information of the sample is missing
check_time <- data.frame(time_within_subject) %>% select(subject, time, rank)

```



# Session info

```{r}
sessionInfo()
```
