---
title: "Time series manipulation"
date: "`r Sys.Date()`"
author:
- name: Leo Lahti
  email: leo.lahti@iki.fi
- name: Yagmur Simsek
  email: yagmur.simsek@hsrw.org
package: 
    miaTime
output: 
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: yes
        toc_depth: 2
        number_sections: true
vignette: >
    %\VignetteIndexEntry{miaTime}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r eval=FALSE, include=FALSE}
library(Cairo)
knitr::opts_chunk$set(cache = FALSE,
                      fig.width = 9,
                      dpi=300,
                      dev = "png",
                      dev.args = list(type = "cairo-png"),
                      message = FALSE,
                      warning = FALSE)
```

```{r loadhide, message=FALSE, warning=FALSE, echo=FALSE}
library(mia)
library(miaTime)
library(dplyr)
library(lubridate)
library(SummarizedExperiment)
```

# Introduction

`miaTime` implements tools for time series manipulation based on the `TreeSummarizedExperiment` [@TSE] data container. Much of the functionality is also applicable to the `SummarizedExperiment` [@SE] data objects. This tutorial shows how to use `miaTime` methods as well as the broader R/Bioconductor ecosystem to manipulate time series data.

Check also the related package [TimeSeriesExperiment](https://github.com/nlhuong/TimeSeriesExperiment).

## Installation

Installing the latest development version in R.

```{r, eval=FALSE}
library(devtools)
devtools::install_github("microbiome/miaTime")
```

Loading the package:

```{r load-packages, message=FALSE, warning=FALSE}
library(miaTime)
```


## Sorting samples 

The _tidySingleCellExperiment_ package provides handy functions for (Tree)SE manipulation.

```{r sort samples, message=FALSE, warning=FALSE}
library("tidySingleCellExperiment")
data(hitchip1006)
tse <- hitchip1006
tse2 <- tse %>% arrange(subject, time)
```


## Storing time information with Period class

`miaTime` utilizes the functions available in the package `lubridate`
to convert time series field to "Period" class object. This gives access to a 
number of readily available [time series manipulation tools](https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html).


Load example data:

```{r}
# Load packages
library(miaTime)
library(lubridate)
library(SummarizedExperiment)

# Load demo data
data(hitchip1006)
tse <- hitchip1006

# Time is given in days in the demo data.
# Convert days to seconds
time_in_seconds <- 60*60*24*colData(tse)[,"time"]
# Convert the time data to period class
Seconds <- as.period(time_in_seconds, unit="sec")
# Check the output
Seconds[1140:1151]
```


## Conversion between time units

The time field in days is now shown in seconds. It can then be
converted to many different units using the lubridate package.

```{r}
Hours <- as.period(Seconds, unit = "hour")
Hours[1140:1151]
```

The updated time information can then be added to the
`SummarizedExperiment` data object as a new `colData` (sample data)
field.


```{r}
colData(tse)$timeSec <- Seconds
colData(tse)
```

## Calculating time differences 

The \link[lubridate]{as.duration} function helps to specify time points as duration.

```{r}
Duration <- as.duration(Seconds)
Duration[1140:1151]
```

The difference between subsequent time points can then be calculated.

```{r}
Timediff <- diff(Duration)
Timediff <- c(NA, Timediff)
Timediff[1140:1151]
```

The time difference from a selected point to the other time points
can be calculated as follows.

```{r}
base <- Hours - Hours[1] #distance from starting point
base[1140:1151]

base_1140 <- Seconds - Seconds[1140]
base_1140[1140:1151]
```

## Time point rank

Rank of the time points can be calculated by `rank` function provided in base R.

```{r}
colData(tse)$rank <- rank(colData(tse)$time)
colData(tse)
```


## Operations per unit

Sometimes we need to operate on time series per unit (subject,
reaction chamber, sampling location, ...).

Add time point rank per subject.

```{r}
library(dplyr)
colData(tse) <- colData(tse) %>%
   as.data.frame() %>%
   group_by(subject) %>%
   mutate(rank = rank(time, ties.method="average")) %>%
   DataFrame()
```


## Subset to baseline samples

```{r}
library(tidySummarizedExperiment)

# Pick samples with time point 0
tse <- hitchip1006 |> filter(time == 0)
# Or: tse <- tse[, tse$time==0]

# Sample with the smallest time point within each subject
colData(tse) <- colData(tse) %>%
   as.data.frame() %>%
   group_by(subject) %>%
   mutate(rank = rank(time, ties.method="average")) %>%
   DataFrame

# Pick the subset including first time point per subject
tse1 <- tse[, tse$rank == 1]
```



# Session info

```{r}
sessionInfo()
```
