---
title: "Advanced time series analysis methods"
date: "`r Sys.Date()`"
package: 
    miaTime
output: 
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: yes
        toc_depth: 2
        number_sections: true
vignette: >
    %\VignetteIndexEntry{miaTime}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false

library(knitr)
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    cache = TRUE,
    message = FALSE,
    warning = FALSE,
    echo = TRUE
)
```
## Overview
This vignette demonstrates the use of three key analytical methods in the 
`miaTime` package: `getStability()`, `getBimodality()`, and 
`getShortTermChange()`. These functions provide essential tools for 
understanding temporal dynamics in microbiome data.


### What these methods do
The three methods address different aspects of temporal microbiome analysis:

- getStability(): Quantifies intermediate stability of microbial taxa with 
respect to a reference point, helping identify taxa with predictable versus 
unpredictable temporal behavior

- getBimodality(): Calculates the coefficient of bimodality to detect taxa that 
exhibit bimodal abundance patterns over time

- getShortTermChange(): Computes growth metrics for short-term changes in 
abundance, including growth rates and rates of change

### Why these methods matter
Understanding temporal dynamics is crucial for microbiome research because:

- Stability assessment helps identify resilient core microbiome members versus 
volatile taxa

- Bimodality detection reveals conditionally rare taxa (CRT) that may have 
important ecological roles

- Short-term change analysis captures rapid microbial responses to perturbations 
or environmental shifts

## Setup and data preparation

```{r}
# Load required packages
library(miaTime)
library(mia)
library(ggplot2)
library(dplyr)

# Load example datasets
data(minimalgut)
data(SilvermanAGutData)

# Basic data exploration and preprocessing
tse <- minimalgut
tse <- transformAssay(tse, method = "rclr")
```

## Stability analysis with `getStability()`
### Basic stability assessment
Purpose: Identify taxa with predictable temporal behavior versus those that 
fluctuate unpredictably.

```{r}
# Calculate stability for single bioreactor
tse_sub <- tse[, tse[["StudyIdentifier"]] == "Bioreactor A"]
stability_results <- getStability(
  tse_sub, 
  assay.type = "rclr", 
  time.col = "Time.hr"
)

kable(head(stability_results), 
      caption = "stability for bioreactor A",
      align = "c")

```

### Advanced stability analysis
Use Case: Compare stability across different systems or conditions.

```{r}
# Multi-group stability analysis with linear model
tse <- addStability(
  tse, 
  assay.type = "rclr", 
  time.col = "Time.hr",
  group = "StudyIdentifier", 
  mode = "lm"
)

# Visualize stability differences
stability_data <- rowData(tse) %>%
  as.data.frame() %>%
  select(contains("stability")) %>%
  tibble::rownames_to_column("Taxa") %>%
  tidyr::gather(key = "Group", value = "Stability", -Taxa) %>%
  mutate(Group = gsub("stability_", "", Group))

# Create stability comparison plot
p_stability_dist <- ggplot(stability_data, aes(x = Stability, fill = Group)) +
  geom_histogram(alpha = 0.7, bins = 30, position = "identity") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Distribution of stability values across groups",
       subtitle = "Positive values indicate predictable behavior",
       x = "Stability value", y = "Abundance") +
  theme_minimal() +
  theme(legend.position = "bottom")
p_stability_dist
```

```{r}
# Box plot comparing stability across groups
p_stability_box <- ggplot(stability_data, aes(x = Group, y = Stability, fill = Group)) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Stability comparison across different systems",
       x = "Study group", y = "Stability value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

p_stability_box
```

```{r}
# Identify most stable and unstable taxa
top_stable <- stability_data %>%
  group_by(Taxa) %>%
  summarize(Mean_Stability = mean(Stability, na.rm = TRUE)) %>%
  arrange(desc(Mean_Stability)) %>%
  slice_head(n = 10) %>%
  mutate(Category = "Most stable")

top_unstable <- stability_data %>%
  group_by(Taxa) %>%
  summarize(Mean_Stability = mean(Stability, na.rm = TRUE)) %>%
  arrange(Mean_Stability) %>%
  slice_head(n = 10) %>%
  mutate(Category = "Most unstable")

stability_extremes <- bind_rows(top_stable, top_unstable)

p_stability_extremes <- ggplot(stability_extremes, 
                              aes(x = reorder(Taxa, Mean_Stability), 
                                  y = Mean_Stability, 
                                  fill = Category)) +
  geom_col() +
  coord_flip() +
  labs(title = "Most stable and unstable Taxa",
       x = "Taxa", y = "Mean stability value") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))

p_stability_extremes

```


### Interpreting stability results

- Positive stability values indicate predictable behavior: larger distance from 
reference correlates with larger changes

- Negative values suggest unstable dynamics: fluctuations are negatively linked 
with abundance level

- Values near zero indicate no clear relationship between reference distance and 
changes

## Bimodality analysis with `getBimodality()`

### Detecting bimodal taxa
Purpose: Identify taxa that switch between high and low abundance states.

```{r}
# Calculate bimodality coefficient
tse_gut <- SilvermanAGutData[, SilvermanAGutData[["Vessel"]] == 1]
tse_gut <- transformAssay(tse_gut, method = "relabundance")

bimodality_results <- getBimodality(tse_gut, assay.type = "relabundance")

# Identify highly bimodal taxa (threshold > 0.95)
bimodal_taxa <- rownames(bimodality_results)[
  which(bimodality_results[[1]] > 0.95)
]

# Bimodality coefficient distribution
bimodality_df <- data.frame(
  Taxa = names(bimodality_results),
  Bimodality = bimodality_results[[1]],
  stringsAsFactors = FALSE
)

p_bimodality_dist <- ggplot(bimodality_df, aes(x = Bimodality)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 0.95, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = 0.95, y = Inf, label = "Bimodal threshold (0.95)", 
           vjust = 1.5, hjust = 1.1, color = "red") +
  labs(title = "Distribution of bimodality coefficients",
       subtitle = "Red line indicates threshold for bimodal classification",
       x = "Bimodality coefficient", y = "Abundance") +
  theme_minimal()

p_bimodality_dist
```


### Conditionally rare taxa (CRT) detection
Use Case: Combine bimodality with abundance to find CRT.

```{r}
# Find abundant taxa (>0.5% relative abundance)
abundant_taxa <- getAbundant(
  tse_gut, 
  assay.type = "relabundance", 
  abundant.th = 0.1/100
)

# Identify CRT as intersection of bimodal and abundant taxa
crt_taxa <- intersect(bimodal_taxa, abundant_taxa)

```

```{r}
# Visualize CRT temporal patterns
if (length(crt_taxa) > 0) {
  # Select top 6 CRT taxa for visualization
  selected_crt <- head(crt_taxa, 6)
  
  # Extract abundance data for CRT taxa
  crt_abundance <- assay(tse_gut, "relabundance")[selected_crt, ] %>%
    as.data.frame() %>%
    tibble::rownames_to_column("Taxa") %>%
    tidyr::gather(key = "SampleID", value = "Abundance", -Taxa) %>%
    left_join(
      colData(tse_gut) %>% 
        as.data.frame() %>% 
        tibble::rownames_to_column("Sample") %>%
        select(SampleID, DAY_ORDER, SampleType),
      by = "SampleID"
    )
  
  p_crt_patterns <- ggplot(crt_abundance, aes(x = DAY_ORDER, 
                                              y = Abundance, color = Taxa)) +
    geom_line(size = 1.2) +
    geom_point(size = 2) +
    facet_wrap(~SampleType, scales = "free_x") +
    labs(title = "Temporal patterns of conditionally rare taxa (CRT)",
         subtitle = "Taxa showing bimodal abundance patterns",
         x = "Time", y = "Relative Abundance") +
    theme_minimal() +
    theme(legend.position = "bottom")
  p_crt_patterns
}

```

```{r}
# Calculate mean abundance for each taxa
mean_abundance <- rowMeans(assay(tse_gut, "relabundance"), na.rm = TRUE)

bimodality_abundance <- data.frame(
  Taxa = rownames(bimodality_results),
  Bimodality = bimodality_results[[1]],
  Mean_Abundance = mean_abundance[rownames(bimodality_results)],
  stringsAsFactors = FALSE
) %>%
  mutate(
    Classification = case_when(
      Bimodality > 0.95 & Mean_Abundance > 0.005 ~ "CRT",
      Bimodality > 0.95 ~ "Bimodal",
      Mean_Abundance > 0.005 ~ "Abundant",
      TRUE ~ "Other"
    )
  )

p_bimodal_abundance <- ggplot(bimodality_abundance, 
                             aes(x = Mean_Abundance, 
                                 y = Bimodality, color = Classification)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 0.005, linetype = "dashed", color = "blue") +
  scale_x_log10() +
  labs(title = "Bimodality vs. Mean abundance",
       subtitle = "CRT taxa are both bimodal and abundant",
       x = "Mean relative abundance (log10)", y = "Bimodality coefficient") +
  theme_minimal() +
  theme(legend.position = "bottom")

p_bimodal_abundance

```


### Biological Interpretation

- High bimodality (>0.95) suggests taxa with switching dynamics

- Combined with abundance, identifies conditionally rare taxa that may have 
keystone roles

- These taxa often respond strongly to environmental changes or host conditions


## Short-term change analysis with `getShortTermChange()`

### Growth rate analysis
Purpose: Quantify rapid microbial responses and growth dynamics.

```{r}
# Calculate short-term changes
tse_change <- transformAssay(tse, method = "relabundance")
change_results <- getShortTermChange(
  tse_change,
  assay.type = "relabundance",
  time.col = "Time.hr",
  group = "StudyIdentifier"
)

# Examine growth metrics
kable(head(change_results), 
      caption = "short-term changes",
      align = "c")

```

### Identifying fast-growing taxa
Use Case: Find taxa with highest growth rates during different phases.

```{r}
change_results <- as.data.frame(change_results)
# Select taxa with highest absolute growth rates
top_growers <- change_results %>%
  filter(!is.na(growth_rate)) %>%
  arrange(desc(abs(growth_rate))) %>%
  slice_head(n = 10)

# Visualize growth trajectories
growth_plot <- ggplot(top_growers, 
                     aes(x = Time.hr, y = growth_rate, color = FeatureID)) +
  geom_line() +
  facet_wrap(~StudyIdentifier, scales = "free") +
  labs(title = "Growth Rate Dynamics of Top Growing Taxa",
       x = "Time (hours)", y = "Growth Rate")

# Visualize growth trajectories for top growers
top_growers <- change_results %>%
  filter(!is.na(growth_rate)) %>%
  group_by(FeatureID) %>%
  summarize(Mean_Growth = mean(abs(growth_rate), na.rm = TRUE)) %>%
  arrange(desc(Mean_Growth)) %>%
  slice_head(n = 10) %>%
  pull(FeatureID)

growth_trajectories <- change_results %>%
  filter(FeatureID %in% top_growers) %>%
  arrange(FeatureID, Time.hr)

p_growth_dynamics <- ggplot(growth_trajectories, 
                           aes(x = Time.hr, y = growth_rate, color = FeatureID)) +
  geom_line(size = 1) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  facet_wrap(~StudyIdentifier, scales = "free") +
  labs(title = "Growth rate dynamics of top growing taxa",
       x = "Time (hours)", y = "Growth rate") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
p_growth_dynamics

```

```{r}
# Create time bins for analysis
growth_binned <- change_results %>%
  filter(!is.na(growth_rate)) %>%
  mutate(
    Time_Bin = cut(Time.hr, 
                   breaks = quantile(Time.hr, probs = seq(0, 1, 0.25), na.rm = TRUE),
                   labels = c("Early", "Mid-Early", "Mid-Late", "Late"),
                   include.lowest = TRUE)
  )

p_growth_distribution <- ggplot(growth_binned, aes(x = Time_Bin, 
                                                   y = growth_rate, 
                                                   fill = Time_Bin)) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Growth rate distribution across time periods",
       x = "Time period", y = "Growth rate") +
  theme_minimal() +
  theme(legend.position = "none") + scale_y_log10()

p_growth_distribution

```

```{r}
# Create heatmap of growth rates for selected taxa
library(pheatmap)

growth_matrix <- change_results %>%
  filter(FeatureID %in% top_growers) %>%
  filter(StudyIdentifier %in% c("Bioreactor A")) %>%
  select(FeatureID, Time.hr, growth_rate) %>%
  tidyr::pivot_wider(names_from = Time.hr, 
              values_from = growth_rate, values_fill = 0) %>%
  tibble::column_to_rownames("FeatureID") %>%
  as.matrix()

# Create heatmap
pheatmap(growth_matrix,
         main = "Growth rate heatmap for top growing taxa in Bioreactor A",
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         scale = "row",
         fontsize = 8)

```


### Logarithmic transformation analysis

```{r}
# Apply log transformation with pseudocount
tse_log <- transformAssay(
  tse_change, 
  assay.type = "relabundance", 
  method = "log10", 
  pseudocount = TRUE
)

# Calculate changes on log scale
log_changes <- getShortTermChange(
  tse_log,
  assay.type = "log10",
  time.col = "Time.hr",
  group = "StudyIdentifier"
)

```

## Integrative analysis
### Combining all three methods
How do stability, bimodality, and growth dynamics relate?

```{r}
tse <- transformAssay(tse, method = "relabundance")
bimodality_results <- getBimodality(tse, assay.type = "relabundance")
# Combine results from all three methods
combined_results <- data.frame(
  Taxa = rownames(tse),
  Stability = stability_results$stability,
  Bimodality = bimodality_results$bimodality,
  stringsAsFactors = FALSE
)

# Add growth rate summaries
growth_summary <- change_results %>%
  group_by(FeatureID) %>%
  summarize(
    Mean_Growth_Rate = mean(abs(growth_rate), na.rm = TRUE),
    Growth_Variability = sd(growth_rate, na.rm = TRUE)
  )

```

```{r}
# Create comprehensive analysis combining all three metrics
library(patchwork)

# Combine results from all three methods
combined_results <- data.frame(
  Taxa = rownames(tse),
  Stability = rowData(tse)$`stability_Bioreactor A`,  # Example for one group
  stringsAsFactors = FALSE
) %>%
  left_join(
    data.frame(
      Taxa = rownames(bimodality_results),
      Bimodality = bimodality_results[[1]],
      stringsAsFactors = FALSE
    ),
    by = "Taxa"
  ) %>%
  left_join(
    change_results %>%
      group_by(FeatureID) %>%
      summarize(
        Mean_Growth_Rate = mean(abs(growth_rate), na.rm = TRUE),
        Growth_Variability = sd(growth_rate, na.rm = TRUE)
      ) %>%
      rename(Taxa = FeatureID),
    by = "Taxa"
  ) %>%
  filter(!is.na(Stability) & !is.na(Bimodality) & !is.na(Mean_Growth_Rate))

# Panel A: Stability vs Bimodality
p_stab_bimod <- ggplot(combined_results, aes(x = Stability, y = Bimodality)) +
  geom_point(alpha = 0.6, size = 2, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "A) Stability vs Bimodality",
       x = "Stability", y = "Bimodality coefficient") +
  theme_minimal()

# Panel B: Stability vs Growth Rate
p_stab_growth <- ggplot(combined_results, aes(x = Stability, 
                                              y = Mean_Growth_Rate)) +
  geom_point(alpha = 0.6, size = 2, color = "forestgreen") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "B) Stability vs Growth rate",
       x = "Stability", y = "Mean Growth Rate") +
  theme_minimal()

# Panel C: Bimodality vs Growth Rate
p_bimod_growth <- ggplot(combined_results, aes(x = Bimodality, 
                                               y = Mean_Growth_Rate)) +
  geom_point(alpha = 0.6, size = 2, color = "purple") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "C) Bimodality vs Growth rate",
       x = "Bimodality coefficient", y = "Mean growth rate") +
  theme_minimal()

# Panel D: 3D relationship summary
p_summary <- ggplot(combined_results, 
                   aes(x = Stability, y = Bimodality, 
                       size = Mean_Growth_Rate)) +
  geom_point(alpha = 0.6, color = "orange") +
  labs(title = "D) Integrated View",
       x = "Stability", y = "Bimodality coefficient",
       size = "Growth Rate") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Combine all panels
combined_plot <- (p_stab_bimod + p_stab_growth) / (p_bimod_growth + p_summary)
combined_plot <- combined_plot + plot_annotation(
  title = "Integrative analysis of microbiome temporal dynamics",
  subtitle = "Relationship between stability, bimodality, and growth dynamics"
)

print(combined_plot)

```

```{r}
# Create ecological classification based on all three metrics
combined_classified <- combined_results %>%
  mutate(
    Ecological_Role = case_when(
      Stability > 0.5 & Bimodality < 0.5 ~ "Core Stable",
      Stability < -0.5 & Bimodality > 0.95 ~ "Conditionally Rare",
      Mean_Growth_Rate > quantile(Mean_Growth_Rate, 0.9, 
                                  na.rm = TRUE) ~ "Rapid Responder",
      Stability > 0 & Bimodality < 0.5 & 
          Mean_Growth_Rate < quantile(Mean_Growth_Rate, 
                                      0.5, na.rm = TRUE) ~ "Persistent",
      TRUE ~ "Other"
    )
  )

p_ecological_roles <- ggplot(combined_classified, 
                            aes(x = Stability, y = Bimodality, 
                                color = Ecological_Role, 
                                size = Mean_Growth_Rate)) +
  geom_point(alpha = 0.7) +
  labs(title = "Ecological classification of microbial taxa",
       subtitle = "Based on stability, bimodality, and growth dynamics",
       x = "Stability", y = "Bimodality coefficient",
       color = "Ecological role", size = "Growth rate") +
  theme_minimal() +
  theme(legend.position = "bottom")
p_ecological_roles
```

```{r}
# Create detailed time series for representative taxa from each category
representative_taxa <- combined_classified %>%
  group_by(Ecological_Role) %>%
  slice_head(n = 2) %>%
  pull(Taxa)

# Extract time series data
time_series_data <- assay(tse, "rclr")[representative_taxa, ] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Taxa") %>%
  tidyr::gather(key = "Sample", value = "Abundance", -Taxa) %>%
  left_join(
    colData(tse) %>% 
      as.data.frame() %>% 
      tibble::rownames_to_column("Sample") %>%
      select(Sample, Time.hr, StudyIdentifier),
    by = "Sample"
  ) %>%
  left_join(
    combined_classified %>% select(Taxa, Ecological_Role),
    by = "Taxa"
  )

p_time_series <- ggplot(time_series_data, 
                       aes(x = Time.hr, y = Abundance, color = Taxa)) +
  geom_line(size = 1) +
  geom_point(size = 1.5) +
  facet_grid(Ecological_Role ~ StudyIdentifier, scales = "free") +
  labs(title = "Time series patterns by ecological role",
       x = "Time (hours)", y = "Abundance (rclr)") +
  theme_minimal() +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 8))
p_time_series
```

```{r}
# Create comprehensive summary table
summary_stats <- combined_classified %>%
  group_by(Ecological_Role) %>%
  summarize(
    Count = n(),
    Mean_Stability = round(mean(Stability, na.rm = TRUE), 3),
    SD_Stability = round(sd(Stability, na.rm = TRUE), 3),
    Mean_Bimodality = round(mean(Bimodality, na.rm = TRUE), 3),
    SD_Bimodality = round(sd(Bimodality, na.rm = TRUE), 3),
    Mean_Growth_Rate = round(mean(Mean_Growth_Rate, na.rm = TRUE), 3),
    SD_Growth_Rate = round(sd(Mean_Growth_Rate, na.rm = TRUE), 3)
  ) %>%
  arrange(desc(Count))

# Display as formatted table
kable(summary_stats, 
      caption = "Summary statistics by ecological role",
      align = "c")

```


### Ecological interpretations

- Stable + low bimodality: core microbiome members

- Unstable + high bimodality: conditionally rare taxa

- High growth rates + high bimodality: rapid responders to environmental changes

The integration of these methods enables a comprehensive understanding of 
microbiome temporal behavior, supporting applications in personalized medicine, 
environmental monitoring, and microbial ecology research.

## Session info

```{r}
#| label: session_info

sessionInfo()
```
